---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr,logspline)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim <- sim |> na.omit()
str(sim)
```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# contrast-coding predictors
contrasts(sim$Sem) <- c(-0.5, 0.5)
contrasts(sim$Phon) <- c(-0.5, 0.5)

# formula
sim_formula <- bf(Distance ~ Sem*Phon) #Sem + Phon + Sem:Phon

# priors
get_prior(sim_formula, sim)

my_priors <- c(
  prior(normal(74, 34.1), class = "Intercept"),
  prior(normal(0, 12), class = "b")
  ) #sd=12, 95% lies between +-24 

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

xmdl_sim <- brm(sim_formula,
                family = gaussian(),
                prior = my_priors,
                data = sim,
                chains = 4, 
                iter = 2000,
                warmup = 1000,
                thin = 1, 
                seed = 42)

```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

pp_check(xmdl_sim, ndraws = 100)

```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

fixef(xmdl_sim)

mcmc_plot(xmdl_sim, type = "trace") # looks good (like a caterpillar)
mcmc_dens_overlay(xmdl_sim, pars = "b_Sem1")
mcmc_dens_overlay(xmdl_sim, pars = "b_Phon1")
mcmc_dens_overlay(xmdl_sim, pars = "b_Sem1:Phon1")
# all look good
xmdl_sim # Chains have converged for all factors (Rhat = 1.00). High on both the Bulk-ESS (> 4600) and the Tail-ESS (> 3100)
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

#contrasts(sim$Sem) 
#contrasts(sim$Phon)

#visualize posteriors

xmdl_sim |> 
  spread_draws(b_Sem1) |> 
  ggplot(aes(x = b_Sem1)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   scale_x_continuous(limits = c(-30, 10)) +
   labs(x = "Semantic = similar",
        y = "") +
   theme_minimal()

xmdl_sim |> 
  spread_draws(b_Phon1) |> 
  ggplot(aes(x = b_Phon1)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   scale_x_continuous(limits = c(-15, 30)) +
   labs(x = "Phonetic = similar",
        y = "") +
   theme_minimal()


xmdl_sim |> 
  spread_draws(`b_Sem1:Phon1`) |> 
  ggplot(aes(x =`b_Sem1:Phon1`)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   scale_x_continuous(limits = c(-30, 30)) +
   labs(x = "Similar phonetic x similar semantic",
        y = "") +
   theme_minimal()

```

```{r}
#calculate posterior probabilities of direction, 

xmdl_sim |> 
  spread_draws(b_Sem1) |> 
  select(b_Sem1) |> 
  p_direction()
```

```{r}
#calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.
library(logspline)

#describe uncertainty

#For the semantic similarity effect (Sem1), the posterior distribution has a mean of -9.04, with a 95% highest probability density interval between -16.68 and -1.22.For the phonological similarity effect (Phon1), the posterior distribution has a mean of 4.99, with a 95% highest probability density interval between -2.63 and 12.62 (inconsistent direction).For the interaction effect (Sem1:Phon1), the posterior distribution has a mean of -3.15 units, with a 95% highest probability density interval between -17.16 and 10.72.

# extract posteriors
posteriors <- xmdl_sim |> 
  spread_draws(b_Sem1) |> 
  select(b_Sem1)
# refit the model only based on the priors
xmdl_sim_priors <- brm(sim_formula,
                  family = gaussian(),
                  prior = my_priors,
                  sample_prior = "only",
                  data = sim,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 42)

priors <- xmdl_sim_priors |> 
  spread_draws(b_Sem1) |> 
  select(b_Sem1)

# run bayes factor against the null value
bayesfactor_parameters(posteriors, priors, direction = "two-sided", null = 0)

```

## (g) Write up your entire analysis:

Posterior distributions of the model indicate that the semantic change from different to similar , the distance between two cities are estimated to decreases by an average of -9.04 \[95% Credible Interval:-16.68 to -1.22\]. The majority of the posterior samples are negative, indicating that positive values are still plausible but less likely (given the data, the model, and priors).

Subsequently, we calculate the Savage-Dickey density ratio [@DickeyLientz1970-The-Weighted-Li; @WagenmakersLodewyckx2010-Bayesian-hypoth; @roettgermeaningful], i.e., we relate the evidence (the probability of the posterior distribution at point-zero) for the model based on the priors only (i.e., before seeing the data) to the evidence for the model based on both priors and likelihood (i.e., after seeing the data) [@WetzelsGrasman2010-An-encompassing]. The Savage-Dickey method lets us assess evidence for and against a null hypothesis using Bayes factor. Here, the Bayes factor indicates the degree by which the mass of the posterior distribution has shifted further away from the point null value (relative to the prior distribution). Our analysis yields a Bayes Factor of 4.80,indicating that the probability of the effect semantic = similar being point null has decreased , suggesting *moderate* evidence for the alternative hypothesis that the effect of Semantic = similar is zero[@lee-wagenmakers_Bayesian_book_2014].

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
